A multi-scale story of the diffusion of a new technology: the web

### 2025-01-03 paper/submission2/tranos2025.tex
ssh problem

\IfFileExists{xurl.sty}{\usepackage{xurl}}{} % add URL line breaks if available
\urlstyle{same} % disable monospaced font for URLs
\hypersetup{
  pdftitle={A multi-scale story of the diffusion of a new technology: the web},
  pdfkeywords={diffusion, technologies, Web, digital},
  colorlinks=true,
  linkcolor={blue},
  filecolor={Maroon},
  citecolor={Blue},
  urlcolor={Blue},
  pdfcreator={LaTeX via pandoc}}

@@

\begin{frontmatter}
\title{A multi-scale story of the diffusion of a new technology: the
web}

@@

Economic geographers were always interested in how new technologies and
innovations diffuse across space and time with the work of
\citet{hagerstrand1968innovation}, which demonstrated the importance of
communications as a mechanism for the diffusion of innovations, being
the most impactful. It was followed by various empirical studies -- see
for instance \citet{ormrod1990} and \citet{iso2005}. However, the torch
of exploring and modelling such diffusion processes had been passed to
other disciplines such as economics, business studies and sociology well
before the `cultural turn' of economic geography
\citep{perkins2005international}. A potential explanation of the lack of
geographical studies exploring the diffusion of new and, more
specifically for this paper, digital technologies across both space and
time can be attributed to the scarcity of relevant and granular enough
data, a problem also highlighted by \citet{iso2005} and
\citet{kemeny2011international}. As \citet{zook2022mapping} pointed,
digital activities are hardly ever captured in official data.

This paper offers such a contribution: an economic geography study
revealing the importance of spatial mechanisms for the diffusion of an
intangible, digital technology that is the Web in the UK at a high level
of spatial granularity during the 1996-2012 period. It does so by
employing a novel source of big data and machine learning algorithms to
model the \emph{active} engagement with Web during that period and tests
how different spatial and temporal mechanisms shaped its diffusion. This
paper also exemplifies how the combination of data sources and
state-of-the-art research methods which escape the traditional social
science domain can offer new lenses to geographical research regarding
the understanding of technological diffusion.

@@

This gap becomes even more prominent if we consider the importance of
technology and its adoption in economic geography and macroeconomic
growth theories. Simply put, differences in technological adoption and
sophistication may lead to uneven development as new technologies can
increase productivity and allow firms to introduce new products
\citep{solow1957technical, aghion1990model, kemeny2011international}.
Linked to the subject of this paper, the Web -- and the active
engagement with this technology that is creating and maintaining instead
of just browsing a website -- can be considered as the underpinning
technology for the transformative process identified by
\citet{capello2024nexus} as the \emph{digital service economy}.
Importantly, the focus here is not the Web in general, but instead
commercial websites. After all, diffusion together with invention and
innovation are considered the pillars of technological change
\citep{das2022diffusion}. Findings from previous studies illustrated
that early engagement with digital technologies such as the Web led to
longstanding positive productivity gains for these regions that hosted
early adopters \citep{tranosuk}. Hence, obtaining a robust understanding
of the spatial diffusion mechanisms is crucial in mapping future
development trajectories.

Another line of motivation has to do with the intangible nature of the
Web and the debate about the footloose nature of digital technologies,
which contrary to several earlier general purpose technologies are
`light in weight' \citep{coyle1999weightless}. Following the seminal
work of \citet{gaspar1998information}, which demonstrated that
technological improvements in telecommunications lead to increased
demand for face-to-face communications and, consequently, enhance the
importance of cities, empirical studies from economic geography and
urban economics explored whether digital technologies substitute or
complement agglomeration externalities and spatial structure
\citep{kolko_death_2000, ioannides2008effect, tranos2021ubiquitous}.
With the exception of \citet{sinai2004geography}, who focused, among
other things, on web content and demonstrated that more such content is
available in larger markets, this strand of literature has been mostly
concerned with the spatiality of the Internet's hard and tangible
infrastructure and shied away from exploring the spatial dimensions of
one of the most distinct software layers of the Internet infrastructure,
the Web.

Technological diffusion, which is by definition an aggregated process,
can be discussed in parallel with individual adoption mechanisms. On the
one hand, \citet{rogers2010diffusion} identifies early adopter of new
technologies as `knowledgeable risk takers' and \citet{griliches1957} as
`profit maximisers' \citep{ding2010modeling}. Such individual agents are
rewarded because of their attitude towards new technologies and
innovations. On the other hand, \citet{perkins2011internet} attribute
diffusion to two processes: (i) epidemic-like mechanisms, which are
governed by distance, proximity and social interactions, and (ii)
economic mechanisms as new innovations are adopted by users as they
become more profitable, valueable and useful.

@@

The data used here depict the active engagement with the Web as they
contain geolocated and time-stamped counts of commercial websites from
the UK. Instead of adopting metrics reflecting the passive engagement
with the Internet such as Internet adoption and speeds
\citep[e.g.][]{blank2018local, destefano2022fuzzy}, this paper observes
yearly counts of geolocated websites, which reflect the creation and
maintenance of websites. It needs to be highlighted here that although
there are various types of underpinning web technologies -- namely
markup, stylesheet, client- and server-side coding languages as well as
different web browsers and other related programming language -- the
focus here is on the diffusion of the Web in its totality. For a more
detailed and cross-country technological diffusion study, see
\citet{PAPAGIANNIDIS2015308}.

@@

\section{Literature review}\label{sec-litreview}

Geographical diffusion is a synthesis of different processes. On the one
hand, a purely spatial or, in other words, contagious processes can be
identified. Adjacency and, more broadly, distance are the key drivers of
diffusion. This perspective draws similarities with epidemics:
innovations just like pathogens spread because of contagion and,
consequently, proximity and exposure \citep{hivner2003facilitating}. On
the other hand, there is a hierarchical processes. According to this,
instead of a horizontal distance-based diffusion mechanism, the top-down
hierarchy of urban systems shapes technological diffusion. In reality,
the synthesis of these two processes better reflects how new
technologies diffuse over space \citep{morrill2020spatial}.

These ideas were firstly introduced by Torsten Hägerstrand and his
thesis entitled `Innovation Diffusion as a Spatial Process'
\citep{hagerstrand1968innovation}. Hägerstrand was the first to identify
diffusion as a geographical process. The starting point was the idea
that diffusion is based on passing information through social networks,
which themselves tend to be defined by geography. Hence, he identified a
`neighbourhood' mechanism of how information, and consequently,
innovation diffuse. He used agricultural innovations to test and model
his ideas using Monte Carlo simulations. Hägerstrand also incorporated
the role of hierarchy and how some phenomena maybe firstly adopted in
larger cities and then diffuse to second tier ones. This is a sequential
instead of a simultaneous process, which resembles the `lead-lag'
spatial acceleration effect in market research
\citep{bento2018time, PERES201091}. Hägerstrand is more widely known
though for highlighting the role time plays in the diffusion of
innovations: an early-pioneering period, a middle fast accelerating
period and a final saturation one \citep{morrill2020spatial}.

The temporal dimension was further explored by Everett Rogers and his
seminal work on `Diffusion of Innovations' \citep{rogers2010diffusion}.
Rogers being a sociologist focused not on the diffusion of innovations
over space and time, but instead on the adoption of new technologies and
innovations by individuals and the individual mechanisms that drive the
decisions behind adoption. He identified five groups of individuals
regarding their adoption speed: innovators, early adopters, early
majority, late majority and laggards. The key mechanism of diffusion and
adoption is communication and how knowledge is transferred within a
social system. All above approaches agree that diffusion and the
cumulative adoption of a new technology are represented by an S-shaped
line plot \citep{grubler1990rise}.

Schmidt's Law empirically illustrates a similar pattern. \emph{Core} and
usually highly agglomerated regions is where new technologies are
invented and commercially deployed \citep{grubler1990rise}. This is
where the first adopters tend to be based. Then, technologies spread to
the \emph{rim} and eventually to the \emph{periphery}. Although adoption
pace might be higher when new technologies finally arrive to the
periphery, the saturation levels there may never reach the ones in the
core because of the lack of infrastructure or other necessary
institutions \citep{leibowicz2016representing}.

\citet{grubler1990rise} effectively summarises the three key spatial
diffusion mechanisms: (i) the cumulative level of adoption follows an
S-shaped pattern just like purely temporal models; (ii) diffusion is
shaped by a hierarchical mechanism in a form of a centrifugal force:
from core to periphery; and (iii) diffusion is also shaped by distance
and proximity. These are the three mechanisms that the empirical
analysis in Section~\ref{sec-results} investigates.

The remaining of this section reviews empirical studies which analysed
the diffusion on new technologies over space and time. Although the
spatial dimension is present in most of the following studies, the level
of spatial detail is always more coarse than the one adopted in this
paper. \citet{beardsell1999spatial} studied the evolution of the
computer industry in 317 US metro areas during the 1977-1992 period
using employment data. Their analysis indicated that the relative size
distribution holds for urban computer employment and also urban
heterogeneity is essential in explaining this distribution. In a recent
study, \citet{bednarz2020pulled} focused on wind turbines and modelled
their spatial diffusion across 402 German regions during 1970-2015.
Their key finding is that local demand than local supply was the main
driving factor. \citet{haller2011determinants} employed firm panel data
for Ireland during 2003-2005 and illustrated that when firms are located
in a region or industry with high share of firms having a website, they
are more likely to have a website too.

At a global scale \citet{perkins2005international} explored whether the
diffusion rate of new technologies is driven by a latecomer advantage
and the engagement with the global economy via foreign direct
investments and trade. Their results illustrate that indeed latecomers
and developing countries experience diffusion of new technologies more
rapidly than early adopters and developed countries. At the same scale,
\citet{perkins2011internet} explored whether the adoption of previous
communication technologies that is mail, telegrams and telephones were
shaped by similar socioeconomic factors as the Internet. Their results
indicated common patterns regarding the drivers behind the adoption of
different communication technologies.

Turning to studies that share more technological and scalar similarities
with this paper, \citet{ding2010modeling} modelled the spatial diffusion
of mobile telecommunications across regions in China. Their analysis
indicated that socioeconomic characteristics are important determinants
of the timing, speed and the level of mobile diffusion within China.
Using data from a Hungarian online social network,
\citet{lengyel2020role} analysed its adoption and the churn at a very
granular spatial level. Their results were in agreement with the early
theoretical and empirical contributions reviewed here: assortativity,
urban scaling and distance are the key drivers of spatial diffusion.
\citet{bakher2013diffusion} compared different the growth curves as a
means to model diffusion of online shopping in Australia during the
1998-2009 period. At a global scale \citet{PAPAGIANNIDIS2015308}
modelled the diffusion of different web technologies and practices
across countries. Using similar but less extensive data as the one used
here, their analysis illustrated how the diffusion of different web
technologies and practices follow an S-shaped pattern as well as the
different diffusion rates of the different technologies and practices.

All in all, this paper tests whether the three distinct spatial
diffusion mechanisms shaped the diffusion of the Web in the UK at very
granular spatial scales from the Web's early stages onwards. The next
section discusses the empirical framework.

@@

To capture the diffusion of the Web, a website density metric is
developed for two different geographical scales: the Local Authority
Districts (LAD) and the Output Areas (OA). The former is an
administrative unit and there are 374 LAD in the UK. The latter is the
smallest census-based geographical unit and there are c.~230,000 of them
in the UK. This methodological choice will allow the mapping of the
diffusion of web technologies and the assessment of the diffusion
mechanisms at these two very different spatial scales.

The counts of websites at these scales are calculated using data from
the Internet Archive\footnote{\href{See\%20https://archive.org/}{https://archive.org/}.}
and, specifically, the JISC UK Web Domain Dataset \citep{ukwebarchive}.
The Internet Archive is one of the most complete and oldest archive of
webpages in the world operating since 1996
\citep{ainsworth2011much, holzmann2016dawn}. It is a web crawler, which
discovers webpages by following the hyperlinks of every webpage it
archives. This dataset, which is curated by the British Library,
contains all the archived webpages from the UK country code top level
domain (ccTLD - .uk) from the 1996--2012 period. In essence, this is a
long list of 2.5 billion URLs of archived webpages including also the
archival timestamp.

Instead of using the whole .uk ccTLD, this paper focuses on its
commercial subset, the .co.uk second level domain (SLD). This choices
decrease the heterogeneity of the web data as such commercial websites
have specific aims: they are used to diffuse information, support online
transactions and share opinions
\citep{THELWALL2000441, blazquez2018big}. Although a UK company can
adopt a generic TLD such as .com and these cases escape the data used
here, such omissions should not affect the validity of the results given
the popularity of the .uk ccTLD \citep{tranosuk}: UK consumers prefer to
visit a .uk website when they are searching for products or services
\citep{hope}; and anecdotal evidence indicates that during the first
half of 2000, three .co.uk domains were registered every minute
\citep{oecd_coms}. Importantly, previous studies illustrated that .co.uk
is the most popular UK SLD \citep{tranosuk}.

The text from these webpages was scanned using a regular expression
(regex) to identify strings of text which resemble UK postcodes and one
fifth of them included a mention to a postcode \citep{BL2013geo}. This
information allows the geolocation of the data and the creation of the
LAD and OA counts.

The data cleaning process includes an aggregation step, through which
the archived webpages are aggregated to the parent websites. This
website reconstruction process lend itself to the construction of
\emph{website} instead of \emph{webpage} density metrics. Websites tend
to represent specific organisations or entities, and, arguably, are more
meaningful observational units than webpages, which can ignore the
upstream dependency of the website they belong to. Based on the
following example, all three webpages are part of the same website
(\url{http://www.website.co.uk}). Only the overall website and not the
nested webpages are considered for the counts as otherwise the density
metrics would have been biased towards large, place specific, websites.

@@

{\centering \includegraphics[width=1\textwidth,height=0.7\textheight]{anonymised_files/figure-pdf/unnamed-chunk-2-1.pdf}

@@

@@

\noindent Equation \ref{eq:rf.generic} is estimated using Random Forest
(RF). This is a popular ML algorithm for both regression and
classification problems \citep{biau2012analysis}. It was introduced by
\citet{breiman2001random} and has become a go-to data science tool. RF
can effectively handle skewed distributions and outliers, model
non-linear relationships, require minimal hyperparameter tuning, exhibit
low sensitivity to these parameters, and have relatively short training
times \citep{Caruana2008, liaw2002classification, yan2020using}. These
attributes match well with the website density data characteristics
including skewness especially for the OA. Also, the large data size
(c.~230k data points for each of the 17 years) calls for fast training
times. Importantly, RF predictions tend to be more accurate than those
from single regression trees and outperform Ordinary Least Squares in
out-of-sample predictions, even with moderate-sized training data and a
small number of predictors
\citep{mullainathan2017machine, athey2019machine, sulaiman2011intelligent, pourebrahim2019trip, biau2012analysis}.

@@

{\centering \includegraphics[width=1\textwidth,height=\textheight]{anonymised_files/figure-pdf/morani-1.pdf}

@@

{\centering \includegraphics[width=1\textwidth,height=0.8\textheight]{anonymised_files/figure-pdf/unnamed-chunk-6-1.pdf}

@@

{\centering \includegraphics[width=1\textwidth,height=\textheight]{anonymised_files/figure-pdf/gini-1.pdf}

@@

\begin{align}
R^2 = 1 - \frac{\sum_{k} (y_{k} - \hat{y_{k}})^2} {\sum_{k} (y_{k} - \overline{y_{k}})^2} \label{eq:rsquared}
\end{align}

\begin{align}
MAE = \frac{1}{N} \sum_{k = 1}^{N} |\hat{y_{k}} - y_{k}| \label{eq:mae}
\end{align}

\begin{align}
RMSE =  \sqrt{\frac{\sum_{k = 1}^{N} (\hat{y_{k}} - y_{k})^2} {N}} \label{eq:rmse}
\end{align}

@@

\begin{longtable}[]{@{}llll@{}}
\caption{Model metrics without data
imputation}\label{tbl-model-metrics-no-correction}\tabularnewline
\toprule\noalign{}
& \(RMSE\) & \(R^{2}\) & \(MAE\) \\
\midrule\noalign{}
\endfirsthead
\toprule\noalign{}
& \(RMSE\) & \(R^{2}\) & \(MAE\) \\
\midrule\noalign{}
\endhead
\bottomrule\noalign{}
\endlastfoot
Local Authorities & 0.032 & 0.810 & 0.019 \\
Output Areas & 5.000 & 0.205 & 1.047 \\
\end{longtable}

@@

{\centering \includegraphics[width=1\textwidth,height=0.6\textheight]{anonymised_files/figure-pdf/varimp-1.pdf}

@@

\setcounter{section}{0}
\renewcommand{\thesection}{\Alph{section}}

\setcounter{table}{0}
\renewcommand{\thetable}{A\arabic{table}}

\setcounter{figure}{0}
\renewcommand{\thefigure}{A\arabic{figure}}

@@

{\centering \includegraphics[width=1\textwidth,height=\textheight]{anonymised_files/figure-pdf/morani10-1.pdf}

@@

{\centering \includegraphics[width=1\textwidth,height=0.8\textheight]{anonymised_files/figure-pdf/unnamed-chunk-10-1.pdf}

@@

{\centering \includegraphics[width=1\textwidth,height=\textheight]{anonymised_files/figure-pdf/gini10-1.pdf}

@@

{\centering \includegraphics[width=1\textwidth,height=0.6\textheight]{anonymised_files/figure-pdf/varimp10-1.pdf}

>>>>

\IfFileExists{xurl.sty}{\usepackage{xurl}}{} % add URL line breaks if available
\urlstyle{same} % disable monospaced font for URLs
\hypersetup{
  pdftitle={A multi-scale story of the diffusion of a new technology: the web},
  pdfauthor={Emmanouil Tranos},
  pdfkeywords={diffusion, technologies, Web, digital},
  colorlinks=true,
  linkcolor={blue},
  filecolor={Maroon},
  citecolor={Blue},
  urlcolor={Blue},
  pdfcreator={LaTeX via pandoc}}

@@

\begin{frontmatter}
\title{A multi-scale story of the diffusion of a new technology: the
web}
\author[1]{Emmanouil Tranos%
\corref{cor1}%
}
 \ead{e.tranos@bristol.ac.uk}

\affiliation[1]{organization={University of
Bristol},postcode={UK},postcodesep={}}

@@

Economic geographers were always interested in how new technologies and
innovations diffuse across space and time with the work of
\citet{hagerstrand1968innovation}, which demonstrated the importance of
communications as a mechanism for the diffusion of innovations, being
the most impactful. It was followed by various empirical studies -- see
for instance \citet{ormrod1990} and \citet{iso2005}. However, the torch
of exploring and modelling such diffusion processes had been passed to
other disciplines such as economics, business studies and sociology well
before the `cultural turn' of economic geography
\citep{perkins2005international, ding2010modeling}. As a consequence,
relevant studies did not investigate diffusion mechanisms at granular
geographical scales to provide local policy insights, for instance,
regarding infrastructural d evelopment and local technology policies. A
potential explanation of the lack of geographical studies exploring the
diffusion of new and, more specifically for this paper, digital
technologies across both space and time can be attributed to the
scarcity of relevant and granular enough data, a problem also
highlighted by \citet{iso2005} and \citet{kemeny2011international}. As
\citet{zook2022mapping} pointed, digital activities are hardly ever
captured in official data.

This paper offers such a contribution: an economic geography study
revealing the importance of spatial mechanisms for the diffusion of an
intangible, digital technology that is the Web in the UK at a high level
of spatial granularity from the very early days until its maturity
(1996-2012). It specifically focuses on the diffusion of commercial
websites. It does so by employing a novel source of big data and machine
learning algorithms to model the \emph{active} engagement with Web --
that is creating and maintaining instead of just browsing a website --
during that period and tests how different spatial and temporal
mechanisms shaped its diffusion. This paper also exemplifies how the
combination of data sources and state-of-the-art research methods which
escape the traditional social science domain can offer new lenses to
geographical research regarding the understanding of technological
diffusion.

@@

This gap becomes even more prominent if we consider the importance of
technology and its adoption in economic geography and macroeconomic
growth theories. Simply put, differences in technological adoption and
sophistication may lead to uneven development as new technologies can
increase productivity and allow firms to introduce new products
\citep{solow1957technical, aghion1990model, kemeny2011international}.
Linked to the subject of this paper, the Web and, more specifically,
commercial websites can be considered as the underpinning technology for
the transformative process identified by \citet{capello2024nexus} as the
\emph{digital service economy}. Importantly, the focus here is not the
Web in general, but instead commercial websites. After all, diffusion
together with invention and innovation are considered the pillars of
technological change \citep{das2022diffusion}. Findings from previous
studies illustrated that early engagement with digital technologies such
as the Web led to longstanding positive productivity gains for these
regions that hosted early adopters \citep{tranosuk}. Hence, obtaining a
robust understanding of the spatial diffusion mechanisms is crucial in
mapping future development trajectories.

Another line of motivation has to do with the very nature of the Web as
a digital technology and how it evolved over time. The digital
geographies field has debunked premature claims about the Internet's
placeless-ness by mapping various of its facets and technological layers
\citep{tranos2013death}, but the lack of relevant long time-series data
prevented researchers from observing digital phenomena overtime. Most of
these studies focused on its hard infrastructure
\citep[e.g.][]{malecki2002economic, moss2000internet}, on Internet
subscribers \citep[e.g.][]{blank2018local}, or on social media users
\citep[e.g.][]{crampton2013beyond}. Exceptions include early work from
\citet{zook2000web}, \citet{zook2000web} and \citet{moss1997tracking},
which utilised domain name registration data and use it to map and
illustrate the spatial footprint of the web content. Pivotal as these
contributions have been in understanding the economic geographies of the
Internet, they were based on cross-sectional, often spatially detailed,
mapping and on shallow time comparisons (e.g.~1993-1998) and did not
observe the evolution and diffusion of the Web until its maturity. As we
know from the relevant literature, different technologies diffuse at
different rates and there is no theory that can explain these
differences \citep{leibowicz2016representing}. Hence, mapping and
modelling the diffusion of the Web among firms can offer useful
insights.

@@

The data used here depict the active engagement with the Web as they
contain geolocated and time-stamped counts of commercial websites from
the UK. Instead of adopting metrics reflecting the passive engagement
with the Internet such as Internet adoption and speeds
\citep[e.g.][]{blank2018local, destefano2022fuzzy}, this paper observes
yearly counts of geolocated, commercial websites, which reflect the
creation and maintenance of such websites. It needs to be highlighted
here that although there are various types of underpinning web
technologies -- namely markup, stylesheet, client- and server-side
coding languages as well as different web browsers and other related
programming language -- the focus here is on the diffusion of the Web as
an overarching technology instead of more specific web technologies. For
a such a detailed and cross-country technological diffusion study, see
\citet{PAPAGIANNIDIS2015308}.

@@

\section{The mechanisms that shape the diffusion of a new
technology}\label{sec-litreview}

The diffusion of new technologies and innovations is driven by three
different mechanisms \citep{grubler1990rise, morrill2020spatial}: (i)
distance and proximity, (ii) urban hierarchies in the form of a
centrifugal forces as new technologies diffuse from core to periphery,
and (iii) over time the cumulative level of adoption follows an S-shaped
pattern. \citet{hagerstrand1968innovation} was the first to identify
diffusion as a geographical process. The starting point was the idea
that diffusion is based on passing information through social networks,
which themselves tend to be geographically bounded. Hence, he identified
a `neighbourhood' mechanism of how information, and consequently,
innovations and new technologies diffuse. This perspective draws
similarities with epidemics: new technologies and innovations, just like
pathogens, spread because of contagion and, consequently, proximity and
exposure \citep{hivner2003facilitating}.

\citet{hagerstrand1968innovation} also emphasised the role of the
top-down hierarchy of urban systems and highlighted that new
technologies and innovations are firstly adopted in larger cities and
then diffuse to second tier ones. This is a sequential instead of a
simultaneous process, which resembles the `lead-lag' spatial
acceleration effect in market research
\citep{bento2018time, PERES201091}. Schmidt's Law empirically
illustrates a similar pattern \citep{grubler1990rise}. Core and highly
agglomerated regions is where new technologies are invented and
commercially deployed. This is where the first adopters tend to be
based. Then, technologies spread to the rim and eventually to the
periphery. Although adoption pace might be higher when new technologies
finally arrive to the periphery, the saturation levels there may never
reach the ones in the core because of the lack of infrastructure or
other necessary institutions \citep{leibowicz2016representing}.

Diffusion is naturally a dynamic process and Hägerstrand recognised the
role time plays in the diffusion of innovations and new technologies: an
early-pioneering period, a middle fast accelerating period and a final
saturation one \citep{morrill2020spatial}. This was further explored by
\citet{rogers2010diffusion}, who identified five types of adopters of
new technologies and innovations: innovators, early adopters, early
majority, late majority and laggards. Hence, plotting the cumulative
adoption of a new technology over time often leads to an S-shaped line
plot \citep{grubler1990rise}.

Although the above processes have been studied in innovation studies,
the distancing of economic geographers from research questions about the
diffusion of new technologies and innovations resulted to a knowledge
gap regarding the explanatory power of these mechanisms at granular,
local scales \citep{ding2010modeling}. As discussed in
Section~\ref{sec-introduction}, such spatially explicit knowledge can
directly inform local technology and infrastructure policies. Indeed,
while the spatial dimension is present in the following studies, the
geographical level of analysis is often too coarse to ascertain the role
of the above mechanisms at the local scale. \citet{beardsell1999spatial}
studied the evolution of the computer industry in 317 US metropolitan
areas during the 1977-1992 period using employment data. Their analysis
indicated that the relative size distribution holds for urban computer
employment and also urban heterogeneity is essential in explaining this
distribution. In a recent study, \citet{bednarz2020pulled} focused on
wind turbines and modelled their spatial diffusion across 402 German
regions during 1970-2015. Their key finding is that local demand than
local supply was the main driving factor. \citet{haller2011determinants}
employed firm panel data for Ireland during 2003-2005 and illustrated
that when firms are located in a region or industry with high share of
firms having a website, they are more likely to have a website too. At a
global scale \citet{perkins2011internet} explored whether the adoption
of previous communication technologies that is mail, telegrams and
telephones were shaped by similar socioeconomic factors as the Internet.
Their results indicated common patterns regarding the drivers behind the
adoption of different communication technologies.
\citet{feldman2015rknowledge} focused on inventions and used patent data
to map how the the rDNA technology diffused among inventors across 366
US metropolitan areas. \citet{ding2010modeling} modelled the spatial
diffusion of mobile telecommunications across the highest level of
administrative division in China (29 regions in total). Their analysis
indicated that socioeconomic characteristics are important determinants
of the timing, speed and the level of mobile diffusion within China.
\citet{bakher2013diffusion} compared different the growth curves as a
means to model the diffusion of online shopping in Australia during the
1998-2009 period, but they did not consider any spatial mechanisms. At a
global scale \citet{PAPAGIANNIDIS2015308} modelled the diffusion of
different web technologies and practices across countries. Using similar
but less extensive data as the one used here, their analysis illustrated
how the diffusion of different web technologies and practices follows an
S-shaped pattern as well as the different diffusion rates of the
different technologies and practices. The most detailed consideration of
the spatial footprint of the diffusion of a new technology comes from
\citet{lengyel2020role}, who analysed the adoption and churn of a
Hungarian online social network. Their results were in agreement with
the early theoretical and empirical contributions reviewed here:
assortativity, urban scaling and distance are the key drivers of spatial
diffusion. Based on the above, the main empirical hypothesis states:

\begin{quote}
H\textsubscript{1}: The three diffusion mechanisms namely distance,
urban hierarchy and the the S-shaped pattern of the cummulative level of
adoption shaped the diffusion of the commercial Web in the UK even at
granular geographical scales.
\end{quote}

Understanding the diffusion of a new technology in multiple, detailed
spatial scales is of interest to economic geography as potential local
economic benefits may occur because of concentration of early adopters
or through technological leapfrogging. On the one hand, early adopters
are often rewarded because of their attitudes towards new technologies
and innovations. \citet{rogers2010diffusion} identifies them as
`knowledgeable risk takers' and \citet{griliches1957} as `profit
maximisers' \citep{ding2010modeling}. On the other hand, latecomers can
adopt and benefit from new technologies that have been developed
elsewhere without incurring the hefty initial R\&D costs
\citep{teece2008firm}. At a global scale
\citet{perkins2005international} explored whether the diffusion rate of
new technologies is driven by a latecomer advantage and the engagement
with the global economy via foreign direct investments and trade. Their
results illustrate that indeed latecomers and developing countries
experience diffusion of new technologies more rapidly than early
adopters and developed countries. At a regional scale, previous research
highlighted the long term and sustained productivity benefit of the
early adopters of web technologies in the UK \citep{tranosuk}. Although
the early adoption and leapfrogging literature has not paid much
attention on cities and regions \citep{yu2018sustainability}, the
potential local economic benefits justify the need to investigate the
dynamics of the web diffusion in different scales in the UK. Hence, the
second empirical hypothesis states:

\begin{quote}
H\textsubscript{2}: The diffusion of the commecrial Web is shaped by
both early adoption and late comer effects.
\end{quote}

These are the two hypotheses are empirically tested in
Section~\ref{sec-results} using the data and methods described in the
next section.

@@

To capture the diffusion of the commercial Web, a website density metric
is developed for two different geographical scales that are discussed
later in this section. The main data are counts of commercial websites
that are calculated using data from the Internet Archive\footnote{\href{See\%20https://archive.org/}{https://archive.org/}.}
and, specifically, the JISC UK Web Domain Dataset \citep{ukwebarchive}.
The Internet Archive is one of the most complete and oldest archive of
webpages in the world operating since 1996
\citep{ainsworth2011much, holzmann2016dawn}. It is a web crawler, which
discovers webpages by following the hyperlinks of every webpage it
archives. This dataset, which is curated by the British Library,
contains all the archived webpages from the UK country code top level
domain (ccTLD - .uk) from the 1996--2012 period. In essence, this is a
long list of 2.5 billion URLs of archived webpages including also the
archival timestamp.

Instead of using the whole .uk ccTLD, this paper focuses on its
commercial subset, the .co.uk second level domain (SLD). This choice is
aligned with the topic of the paper -- the diffusion of the Web for
commercial functions -- and decreases the heterogeneity of the web data
as such commercial websites have specific aims: they are used to diffuse
information, support online transactions and share opinions
\citep{THELWALL2000441, blazquez2018big}. Although a UK company can
adopt a generic TLD such as .com and these cases escape the data used
here, such omissions should not affect the validity of the results given
the popularity of the .uk ccTLD \citep{tranosuk}: UK consumers prefer to
visit a .uk website when they are searching for products or services
\citep{hope}; and anecdotal evidence indicates that during the first
half of 2000, three .co.uk domains were registered every minute
\citep{oecd_coms}. Importantly, previous studies illustrated that .co.uk
is the most popular UK SLD \citep{tranosuk}. Henceforward any reference
to websites will refer to commercial -- .co.uk -- websites.

The text from these webpages was scanned using a regular expression
(regex) to identify strings of text which resemble UK postcodes and one
fifth of them included a mention to a postcode \citep{BL2013geo}. This
information allows the geolocation of the data and the creation of the
LAD and OA counts. To prevent false positives, the resulted postcodes
were matched against official postcode lookup files \citep{ons}.

The data cleaning process includes an aggregation step, through which
the archived webpages are aggregated to the parent websites. This
website reconstruction process lend itself to the construction of
\emph{website} instead of \emph{webpage} density metrics. Websites tend
to represent specific organisations or entities, and, arguably, are more
meaningful observational units than webpages, which can ignore the
upstream dependency of the website they belong to. Based on the
following example, all three webpages are part of the same website
(\url{http://www.website.co.uk}). Only the overall website and not the
nested webpages are considered for the counts as otherwise the density
metrics would have been biased towards large, place specific, websites.
It is important to note here that websites commonly contain place
references, with previous research estimating this share to be around
70\% \citep{hill2009georeferencing}.

@@

{\centering \includegraphics[width=1\textwidth,height=0.7\textheight]{tranos2025_files/figure-pdf/unnamed-chunk-2-1.pdf}

@@

the Local Authority Districts (LAD) and the Output Areas (OA). The
former is an administrative unit and there are 374 LAD in the UK. The
latter is the smallest census-based geographical unit and there are
c.~230,000 of them in the UK. This methodological choice will allow the
mapping of the diffusion of web technologies and the assessment of the
diffusion mechanisms at these two very different spatial scales.

@@

\noindent Equation \ref{eq:rf.generic} is estimated using Random Forest
(RF). This is a popular ML algorithm for both regression and
classification problems \citep{biau2012analysis}. It was introduced by
\citet{breiman2001random} and has become a go-to data science tool. RF
can effectively handle skewed distributions and outliers, model
non-linear relationships, require minimal hyperparameter tuning, exhibit
low sensitivity to these parameters, and have relatively short training
times \citep{Caruana2008, liaw2002classification, yan2020using}. These
attributes match well with the website density data characteristics
including skewness especially for the OA. Also, the large data size (c.
230k data points for each of the 17 years) calls for fast training
times. Importantly, RF predictions tend to be more accurate than those
from single regression trees and outperform Ordinary Least Squares in
out-of-sample predictions, even with moderate-sized training data and a
small number of predictors
\citep{mullainathan2017machine, athey2019machine, sulaiman2011intelligent, pourebrahim2019trip, biau2012analysis}.

@@

{\centering \includegraphics[width=1\textwidth,height=\textheight]{tranos2025_files/figure-pdf/morani-1.pdf}

@@

{\centering \includegraphics[width=1\textwidth,height=0.8\textheight]{tranos2025_files/figure-pdf/unnamed-chunk-6-1.pdf}

@@

{\centering \includegraphics[width=1\textwidth,height=\textheight]{tranos2025_files/figure-pdf/gini-1.pdf}

@@

\begin{align}
R^2 = 1 - \frac{\sum_{k} (y_{k} - \hat{y_{k}})^2} {\sum_{k} (y_{k} - \overline{y_{k}})^2} \label{eq:rsquared}
\end{align} \begin{align}
MAE = \frac{1}{N} \sum_{k = 1}^{N} |\hat{y_{k}} - y_{k}| \label{eq:mae}
\end{align} \begin{align}
RMSE =  \sqrt{\frac{\sum_{k = 1}^{N} (\hat{y_{k}} - y_{k})^2} {N}} \label{eq:rmse}
\end{align}

@@

\begin{longtable}[]{@{}llll@{}}
\caption{Model metrics without data imputation
}\label{tbl-model-metrics-no-correction}\tabularnewline
\toprule\noalign{}
& \(RMSE\) & \(R^{2}\) & \(MAE\) \\
\midrule\noalign{}
\endfirsthead
\toprule\noalign{}
& \(RMSE\) & \(R^{2}\) & \(MAE\) \\
\midrule\noalign{}
\endhead
\bottomrule\noalign{}
\endlastfoot
Local Authorities & 0.032 & 0.810 & 0.019 \\
Output Areas & 5.000 & 0.205 & 1.047 \\
\end{longtable}

@@

{\centering \includegraphics[width=1\textwidth,height=0.6\textheight]{tranos2025_files/figure-pdf/varimp-1.pdf}

@@

\setcounter{section}{0}
\renewcommand{\thesection}{\Alph{section}}
\setcounter{table}{0}
\renewcommand{\thetable}{A\arabic{table}}
\setcounter{figure}{0}
\renewcommand{\thefigure}{A\arabic{figure}}

@@

{\centering \includegraphics[width=1\textwidth,height=\textheight]{tranos2025_files/figure-pdf/morani10-1.pdf}

@@

{\centering \includegraphics[width=1\textwidth,height=0.8\textheight]{tranos2025_files/figure-pdf/unnamed-chunk-10-1.pdf}

@@

{\centering \includegraphics[width=1\textwidth,height=\textheight]{tranos2025_files/figure-pdf/gini10-1.pdf}

@@

{\centering \includegraphics[width=1\textwidth,height=0.6\textheight]{tranos2025_files/figure-pdf/varimp10-1.pdf}
